*lancer*: an R package for linearity assessment and visualisation of multiple curves

### 2023-01-21 Manuscript/manuscript.qmd
Update scripts based on teams suggestions.

Linearity assessment is an important quality control to verify if the quantitative acquisition range of an instrument or experimental method is reliable. It is applied in many fields in analytical science such as calibration/dilution studies (@rodríguez1993 and @sands2021) and assay development (@ross2003 and @hsieh2008). After analysis, curves are plotted individually for each analyte of interest with a Pearson Correlation Coefficient value for each plotted curve using general-purpose software like Excel. A non-linear curve suggests possible human, systematic or analytical errors that need to be addressed. Examples of these errors are incorrect amount of analyte, imprecise pipette calibration, expired reagents or equipment malfunction.

However, as instrument or experimental method are becoming more high throughput, many measurements can be done in a short time. Having the analyst to individually plot numerous curves to check for linearity is time-consuming. Furthermore, @sonnergaard2006 warns that the Pearson Correlation Coefficient is an ineffective standalone numeric parameter to estimate linearity. While researchers have created other metrics for linearity evaluation, these metrics are rarely implemented in most general-purpose software.

The R package, *lancer* addresses these issues by assisting analysts, to plot curves from many experiments easily with additional metrics, other than the Pearson Correlation Coefficient, that better describe the curve's shape. It also provides an interactive viewer for analysts to group, filter and sort the curves, allowing them to examine problematic ones, such as saturated curves.

@@

Using response curves in metabolomic/lipidomic studies as an example, Supplementary Figure 1 depicts the workflow of *lancer*. The workflow starts with two tables:

*    Curve Signal Data, containing response signals (y-axis for response curve) for each sample and curve. 
*    Curve Batch Annotation, containing curve related information, such as concentration (x-axis for response curve) and response curve batches if any.

@@

Next, summary statistics are calculated via `summarise_curve_table` for each curve. Besides the Pearson Correlation Coefficient, one additional calculation is the Mandel's Fitting Test ($F_{stats}$ in @eq-mandel-test) described by @andrade2013. A low $p$ value from the $F$ test gives sufficient evidence that a quadratic model fits better than a linear model, indicating a non-linear curve.

$$
\begin{aligned}
&F_{stats} = \frac{(n-2) \ \mathrm S_{lin}^2 \ - \ (n-3) \ \mathrm S_{quad}^2}{\mathrm S_{quad}^2} \sim \ F(1,n-3)
\\
\\
&\text{where}
\\
&\mathrm S_{lin}^2 = \frac{1}{n-2} \sum_{i=1}^{n} ( y_{i,lin} - y_{i,true})^2 
\\
&\mathrm S_{quad}^2 = \frac{1}{n-3} \sum_{i=1}^{n} ( y_{i,quad} - y_{i,true})^2
\\ 
&y_{i,lin} \ \text{is the linear model predicted y-axis value.}
\\ 
&y_{i,quad} \ \text{is the quadratic model predicted y-axis value.}
\\ 
&y_{i,true} \ \text{is the true y-axis value.}
\\ 
&n \ \text{is the number of data points.}
\end{aligned}
$$ {#eq-mandel-test} Another calculation is Percent Residual Accuracy ($\%RA$ in @eq-pra) from @logue2018. Ranging from $-\infty$ to $100$, a linear curve gives a $\%RA$ value close to $100$.

@@

The software also calculates the concavity of the fitted quadratic model to identify if the curve is dominantly non-linear at high (concavity $<0$) or low (concavity $>0$) concentrations.

Supplementary Figure 2 gives the summary statistics of three manually generated curves: A linear curve and two curves with a plateau at high concentrations (denoted as saturation regime curve) and low concentrations (denoted as noise regime curve) respectively. The corresponding Pearson Correlation Coefficient values (`r_corr`) are $\ge0.9$ ($0.99$, $0.95$ and $0.98$ respectively), even for the non-linear curves. However, both non-linear curves give a much lower Mandel's Fitting Test $p$ values (`mandel_p_val`) ($1.66 * 10^{-4}$ and $2.56 * 10^{-3}$ respectively vs $0.38$). Likewise, the Percent Residual Accuracy values (`pra_linear`) are much lower in the non-linear curves compared to the linear curve ($62.30$ and $74.69$ respectively vs $94.32$).

@@

After calculating the summary statistics for each curve, *lancer* uses the function `evaluate_linearity` to group the curves according to the workflows proposed in @fig-linearity. Workflow 1 uses the Pearson Correlation Coefficient and Percent Residual Accuracy to determine if the curve is linear (labelled as Good Linearity) or not (labelled as Poor Linearity). Workflow 2 goes one step further, using the Mandel's Fitting Test and the fitted quadratic model's concavity to check if the non-linear curve plateaus at low (labeled as limit of detection) or high (labelled as saturation) concentrations. Non-linear curves that do not follow these trends are labelled as Poor Linearity.

@@

A benchmark workflow using only Pearson Correlation Coefficient value of $0.8$ is compared with Workflow 2 on simulated data sets of `r nrow(linear_curve_summary)` linear curves (labelled as Linear), curves that plateau at low (labelled as Noise Regime) or high (labelled as Saturation Regime) concentrations. Supplementary Figure 3 shows that Workflow 2 better identifies the saturation and noise regime curves than the benchmark workflow. While Workflow 2 correctly classifies less linear curves than the benchmark workflow, its percentage of correctly classified linear curves `r inline_text(simulation_results, variable = wf2_group, level = "Good Linearity", column = "Good Linearity")` is high. See <https://lancer-simulation.netlify.app> for report details. While the threshold values of Pearson Correlation Coefficient and Percent Residual Accuracy are based on the interpretation of @y.h.chan2003 and @logue2018, respectively, they remain subjective and arbitrary. Nevertheless, *lancer* allows optimization of these threshold values according to the analyst's preference.

Although *lancer* can export the results in Excel or pdf, they may be too complex for meaningful interpretation. @fig-visualisation shows a HTML folder, exported by *lancer*, such that clicking on the `index.html` file inside the folder will open an interactive trellis plot that can group, filter and sort curves. This allows room for exploratory data analysis, such as identifying curves with linearity issues or understanding the effects of changing the Pearson Correlation Coefficient threshold to another value. Such information is hard to achieve with the Excel and pdf files. An example of an interactive viewer created by *lancer* can be viewed at <https://lancer-interactive-example.netlify.app>

@@

To check if an instrument or experimental method is reliable, it is crucial to check for linearity. However, there are few software that can do this efficiently at a high throughput setting. R package, *lancer*, rectifies this with functions that plots
many curves quickly and reports alternative curve summary statistics, other than the Pearson Correlation Coefficient, to better describe the shape of curves. It also provides an interactive trellis plot for exploratory data analysis. It is available on GitHub <https://github.com/SLINGhub/lancer> while the documentation and tutorials can be accessed from <https://slinghub.github.io/lancer>.

>>>>

Linearity assessment is an important performance test of an analytical instrument or method when a linear response is assumed. It is applied in many fields in analytical sciences, such as assay development (@ross2003 and @hsieh2008), calibration/dilution studies (@rodríguez1993 and @sands2021) and laboratory tests. After analysis, curves representing response values versus concentration are plotted for each analyte of interest. As the accuracy of an analytical method is linked to linearity, a non-linear behaviour must be recognized and addressed accordingly. Visual inspection of curves is still useful but it must be accompanied by statistical tests as decision parameters.

To individually plot numerous curves and check for linearity is time-consuming. Furthermore, @sonnergaard2006 suggests that the Pearson Correlation Coefficient, a commonly used test, is an ineffective standalone numeric parameter to estimate linearity. While researchers like @vanloco2002, @sanchez2021 and @logue2018 have indicated other metrics for linearity evaluation, these metrics are generally not implemented in most software.

The R package *lancer* addresses these issues by assisting analysts to plot curves from many experiments, with additional metrics that better describe the characteristics of the curve. It also provides users with an interactive viewer to group, filter and sort the curves, allowing the examination of problematic cases, such as the ones generated by saturated signals.

@@

Using response curves in metabolomic/lipidomic studies from @croixmarie2009 and @sands2021 as an example, Supplementary Figure 1 depicts the workflow of *lancer*. The workflow starts with two tables:

*    Curve Batch Annotation, describing the curve(s), such as concentration (x-axis) and response curve batches, if any.
*    Curve Signal Data, containing response values (y-axis) for each sample and curve.

@@

Next, summary statistics are calculated via `summarise_curve_table` for each curve. Besides the Pearson Correlation Coefficient, an additional statistical test is the Mandel’s Fitting Test ($F_{stats}$ in @eq-mandel-test) as described by @andrade2013. The $F$ test can give evidence that a quadratic model fits better than a linear model, indicating a non-linear curve.

$$
\begin{aligned}
&F_{stats} = \frac{(n-2) \ \mathrm S_{lin}^2 \ - \ (n-3) \ \mathrm S_{quad}^2}{\mathrm S_{quad}^2} \sim \ F(1,n-3)
\\
\\
&\text{where}
\\
&\mathrm S_{lin}^2 = \frac{1}{n-2} \sum_{i=1}^{n} ( y_{i,lin} - y_{i,true})^2 
\\
&\mathrm S_{quad}^2 = \frac{1}{n-3} \sum_{i=1}^{n} ( y_{i,quad} - y_{i,true})^2
\\ 
&y_{i,lin} \ \text{is the linear model predicted y-axis value.}
\\ 
&y_{i,quad} \ \text{is the quadratic model predicted y-axis value.}
\\ 
&y_{i,true} \ \text{is the true y-axis value.}
\\ 
&n \ \text{is the number of data points.}
\end{aligned}
$$ {#eq-mandel-test} Another statistical test is Percent Residual Accuracy ($\%RA$ in @eq-pra) from @logue2018. Ranging from $-\infty$ to $100$, according to which a perfectly linear curve gives a $\%RA$ value of $100$.

@@

The software also calculates the concavity of the fitted quadratic model, to identify if the curve is dominantly non-linear at high (concavity $<0$) or low (concavity $>0$) concentrations.

Supplementary Figure 2 gives the summary statistics of three simulated curves (as examples), where one curve is linear and two curves show a plateau at high concentrations (denoted as saturation regime curve) and low concentrations (denoted as noise regime curve), respectively. The corresponding Pearson Correlation Coefficient values (`r_corr`) are $\ge0.9$ ($0.99$, $0.95$ and $0.98$ respectively) even for the non-linear curves. However, non-linear curves are detected by much lower Mandel’s Fitting Test $p$ values (`mandel_p_val`) ($1.66 * 10^{-4}$ and $2.56 * 10^{-3}$ respectively vs $0.38$ for the linear curve) and Percent Residual Accuracy values (`pra_linear`) ($62.30$ and $74.69$ respectively vs $94.32$ for the linear curve).

@@

After calculating the summary statistics for each curve, *lancer* uses the function `evaluate_linearity` to group the curves according to the workflows proposed in @fig-linearity. Workflow 1 uses the Pearson Correlation Coefficient and Percent Residual Accuracy to determine if the curve is linear (labelled as Good Linearity) or not (labelled as Poor Linearity). Workflow 2 goes one step further, using the Mandel’s Fitting Test and the fitted quadratic model’s concavity to check if a non-linear curve plateaus at low (labeled as Noise Regime) or high (labelled as Saturation Regime) concentrations. Non-linear curves that do not follow these trends are labelled as Poor Linearity.

@@

A benchmark workflow using only Pearson Correlation Coefficient value of $0.8$ is compared with Workflow 2 on simulated data sets of `r nrow(linear_curve_summary)` linear curves (labelled as Linear), curves that plateau at low (labelled as Noise Regime) or high (labelled as Saturation Regime) concentrations. Supplementary Figure 3 shows that Workflow 2 better identifies the saturation and noise regime curves than the benchmark workflow. While Workflow 2 correctly classifies a lower number of linear curves than the benchmark workflow, its percentage of correctly classified linear curves, `r inline_text(simulation_results, variable = wf2_group, level = "Good Linearity", column = "Good Linearity")`, is high. See <https://lancer-simulation.netlify.app> for report details. While the threshold values of Pearson Correlation Coefficient and Percent Residual Accuracy are based on the interpretation of @y.h.chan2003 and @logue2018, they remain subjective and arbitrary. Nevertheless, *lancer* allows optimization of these threshold values according to the analyst’s preference.

Although *lancer* can export the results in Excel or pdf, an interactive user interface can generate a better overview of the data. @fig-visualisation shows a HTML folder, exported by *lancer*, such that clicking on the index.html file inside the folder will open an interactive trellis plot that can group, filter and sort curves. This allows room for exploratory data analysis, such as identifying curves with linearity issues or understanding the effects of changing the Pearson Correlation Coefficient threshold to another value. Such information is hard to achieve with other common software. An example of an interactive viewer created by *lancer* can be found at <https://lancer-interactive-example.netlify.app>

@@

Linearity is one of the most important parameters of an analytical method to be evaluated. Our R package, *lancer*, can estimate linearity efficiently in high-throughput settings and with functions that plot many curves quickly, reporting curve summary statistics to better describe the shape of the datasets. It also provides an interactive trellis plot for exploratory data analysis. It is available on GitHub <https://github.com/SLINGhub/lancer> while the documentation and tutorials can be accessed from <https://slinghub.github.io/lancer>.
